---
title: "Analysis of Vaccine Efficacy for BNT162b2 Vaccine from Pfizer and BioNTech"
author: "Jaime Li, Hannah Ryu, Kevin Zhao"
date: "Winter 2025"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(fastR2)
library(ggplot2)
```

# 1) Abstract 
    In this paper, we applied Likelihood and Bayesian methods of statistical analysis to analyse our parameter of interest, the vaccine efficacy rate of the BNT162b2 vaccine for the COVID-19 virus, denoted $\psi$ for this paper. The experimental data is from Pfizer and BioNTech, the producers of the vaccine. Vaccine efficacy is the percentage decrease in the chance of being infected given the vaccine compared to the placebo group and the FDA requires a vaccine efficacy rate of at least 30% for vaccines to be approved for general use. We used the maximum likelihood estimate, likelihood ratio p-value, Wald confidence interval, and a bootstrapped interval for the likelihood analysis. We approached the data using a Bayesian framework as well, computing a Bayesian credible interval and hypothesis testing. Using both methods, we found significant evidence to support that the vaccine efficacy exceeds the 30% threshold, meaning that the BNT162b2 vaccine would be approved by the FDA for general use.

# 2) Keywords
BNT162b2 Vaccine, COVID-19 Vaccine, Vaccine Efficacy, Maximum Likelihood Estimation, Bayesian Estimation


# 3) Introduction 
  COVID-19 is a contagious disease caused by the coronavirus SARS-CoV-2. Several COVID-19 vaccines have been approved and distributed in various countries, many of which have initiated mass vaccination campaigns. In December 2020, Pfizer and BioNTech successfully obtained a US FDA Emergency Use Authorization (EUA) to begin distributing their two-dose vaccine, BNT162b2. BNT162b2 utilizes mRNA technology to stimulate an immune response. About 649 million doses of the Pfizer–BioNTech COVID-19 vaccine, including about 55 million doses in children and adolescents (below 18 years of age) were administered in the EU/EEA from authorization to 26 June 2022. To this day, there are still some people still question the effectiveness of the vaccine. The FDA requires at least 30% vaccine efficacy for a vaccine to be approved. Vaccine efficacy is the percentage reduction of the risk of getting COVID-19 give the vaccine compared to the placebo group.
The experiment is a placebo-controlled, double-blinded randomized trial, enrolling participants aged 16 and older. Participants were randomly assigned in a 1:1 ratio to receive either two doses of BNT162b2 or a placebo, administered 21 days apart. The primary efficacy endpoint was laboratory-confirmed COVID-19 infection, with a total of 170 cases observed across the study groups. 

\begin{table}[h!]
\centering
\begin{tabular}{c|c|c}
\textbf{Group} & \textbf{Cases} & \textbf{Sample Size} \\ \hline
BNT162b2 & 8 & 17,411 \\ 
Placebo & 162 & 17,511 \\ \hline
Total & 170 & 34,922 \\ 
\end{tabular}
\caption{Vaccine Efficacy against COVID-19 at least 7 days
after second dose in patients without evidence of infection, from Pfizer and BioNTech}
\end{table}

```{r echo=FALSE}
groups <- c("BNT162b2", "Placebo")
cases <- c(8, 162)
sample_sizes <- c(17411, 17511)

df <- data.frame(
  Group = rep(groups, each = 2),
  Category = rep(c("Infected Cases", "No Infection"), times = 2),
  Percentage = c(800/17411, (1-8/17411)*100, (162/17511)*100, (1-162/17511)*100)
)

# Create the segmented bar plot
ggplot(df, aes(x = Group, y = Percentage, fill = Category)) +
  geom_bar(stat = "identity", position = "stack") +
  ggtitle("Percentage of COVID-19 Cases in Vaccine and Placebo Groups") +
  xlab("Group") +
  ylab("Percentage") +
  theme_minimal()
```
Given the significance of this vaccine in controlling the pandemic, our report aims to analyze the clinical trial data using both Bayesian and Frequentist statistical methods. Statistical models assessing vaccine efficacy play a crucial role in guiding public health decisions. The evaluation of Pfizer’s clinical trial data through alternative statistical frameworks help reinforce confidence in the robustness of vaccine efficacy estimates and their implications for global vaccination strategies.
The goal of our report is to critically assess the statistical modeling in clinical trials and to understand how different methods may influence public health decision-making.
```{r label="important_R_code", eval=TRUE, echo=FALSE}
# You can reference your code in the appendix (sample here).
```

# 4) Statisical Methods 
In this study, we applied two main statistical methods, Likelihood and Bayesian inference, to analyse our parameter of interest. Our parameter of interest is the vaccine efficacy rate of the BNT162b2 vaccine for the COVID-19 virus, denoted $\psi$ for this paper. Vaccine efficacy is the percentage decrease in the chance of being infected by COVID-19 given the vaccine compared to the placebo group. The FDA requires an efficacy rate of at least 30% before vaccines can be authorized for use, thus the goal of this analysis is to determine if the efficacy rate of the BNT162b2 vaccine sufficiently meets this requirement. Our data is from Pfizer and BioNTech, the producers of the BNT162b2 vaccine.

## 4.1 Likelihood Inference  
Let $n$ be the total number of patients with COVID-19 (from the data, $n$ is equal to 170)\footnote{See Table 1 from the Introduction section.}. Let $T$ be a binomial random variable expressing the number of patients that have COVID-19 that are from the vaccine group.
$$ T \sim Binom(n = 170, \pi)$$

Where $\pi$ denotes the probability of an infected patient being from the vaccine group. Then: $$\pi = P(\text{Vaccine|COVID-19 Infection}) = \frac{\pi_v \cdot n_v}{\pi_v \cdot n_v - \pi_p\cdot n_p}$$
Where $\pi_v$ and $\pi_p$ are the probability of a patient being from the vaccine or placebo group, respectively. Additionally, $n_v$ and $n_p$ are the number of infected patients from the vaccine and placebo group, respectively. Since $n_v \approx n_p$, the randomization is 1:1, meaning we can express the function as:
$$\pi =  \frac{\pi_v }{\pi_v  - \pi_p}$$
Our parameter of interest, the vaccine efficacy, is given by $\psi = 1 - \frac{\pi_v}{\pi_p} = \frac{\pi_p - \pi_v}{\pi_p}$. Note that the range of $\psi = (-\infty, 1)$. Our efficacy formula is also given in terms of $\pi$: $$\psi = \frac{1-2\pi}{1-\pi}$$
Note that $\pi$ can then be written in terms of $\psi$.\footnote{See appendix 8.1 for calculations.}
\begin{align*}
\pi &= \frac{\psi - 1}{\psi - 2}
\end{align*} 

The likelihood function of $\psi$ can be expressed using the pdf of $T$, given that $\pi = \frac{\psi - 1}{\psi - 2}$. 

$$f(T = t) = L(\psi) = \binom{n}{t} \left(\frac{\psi - 1}{\psi - 2}\right)^t \left( 1 - \frac{\psi - 1}{\psi - 2}\right)^{n-t}$$
Where $t$ is the observed number of patients in the vaccine group and $n$ is the number of COVID-19 cases. Then the log-likelihood function:
$$\ell(\psi) = ln\left(\binom{n}{t}\right) + t\:ln\left({\psi - 1}\right) - t \: ln({\psi - 2}) - ({n-t}) \: ln \left( { 2 - \psi}\right)$$

In order to identify the possible maximums of our log-likelihood function, we took the first derivative of our log-likelihood with repects to the parameter of interest, $\psi$, and found the root(s) of the equation.
\begin{align*}
\frac{d}{d\psi} \ell(\psi) &=  \frac{t}{{\psi - 1}} - \frac{t}{{\psi - 2}} + \frac{n-t}{{ 2 - \psi}}\\
&= \frac{t}{{\psi - 1}} + \frac{t}{{ 2 - \psi}} + \frac{n-t}{{ 2 - \psi}}\\
&= \frac{t}{{\psi - 1}} + \frac{t}{{ 2 - \psi}} + \frac{n-t}{{ 2 - \psi}}\\
&= \frac{t}{{\psi - 1}} + \frac{n}{{ 2 - \psi}}
\end{align*}
\begin{align*}
\\
\frac{t}{{\psi - 1}} + \frac{n}{{ 2 - \psi}} = 0\\
\frac{-t}{{\psi - 1}} = \frac{n}{{ 2 - \psi}}\\
n\psi - n = -2t + t\psi\\
n\psi - t\psi = -2t + n\\
\psi = \frac{n - 2t}{n - t}
\end{align*}

This value is a possible maximum of our log-likelihood function. The second derivative test allows us to test the concavity of the function using our given data where $n = 170$ and $t = T_{obs} = 8$: 
\begin{align*}
\frac{d^2}{d\psi^2} \ell(\psi) &=   \frac{-t}{(\psi - 1)^2} + \frac{n}{{ (2 - \psi)^2}}\\\\
\frac{d^2}{d\psi^2} \ell\left(\psi = \frac{n - 2t}{n - t}\right) &= \frac{d^2}{d\psi^2} \ell \left(\frac{170 - 2(8)}{170 - 8} \right) =   \frac{-8}{\left(\left(\frac{154}{162}\right) - 1\right)^2} + \frac{170}{{ \left(2 - \left(\frac{154}{162}\right)\right)^2}} = -3126.124
\end{align*}
Thus our critical point is a local maximum. Thus:
$$\hat{\psi}^{mle}_0 = \frac{n - 2T}{n - T}$$

**The Large Sample Confidence Interval**  
Based on Theorem 12.1 \footnote{ Slide deck "likelihood-inference" from Stat 342, Grove, Ranjini}, the 95% large confidence interval of $\hat{\psi}^{mle}$, under certain regularity conditions, is equal to:
$$\hat{\psi}^{mle}_0 \pm z_{0.975} \sqrt{\frac{1}{n\:I(\hat{\psi}^{mle}_0)}}$$
Where $I(\hat{\psi}^{mle})$ is the expected Fisher Information, equal to:
$$I(\hat{\psi}^{mle}_0) = E\left[\frac{-d^2}{d\psi^2} ln(f_{\psi}(t))\right]_{\psi = \hat{\psi}^{mle}_0}  = \frac{n\pi}{\psi - 1} - \frac{n}{2-\psi} = \frac{t}{\psi - 1} - \frac{n}{2-\psi}$$
Where $\pi$ is the rate of vaccine patients given that $\psi = \hat{\psi}^{mle}_0$. Our large sample confidence interval is then:
$$\hat{\psi}^{mle}_0 \pm z_{0.975} \sqrt{\frac{1}{\frac{t}{\hat{\psi}^{mle}_0 - 1} - \frac{n}{2-\hat{\psi}^{mle}_0}}}$$ 
The key regularity and size assumptions we must consider are\footnote{Slide deck "likelihood-inference" from Stat 342, Grove, Ranjini}:

1) The possible values of $t$ are not dependent on the parameter $\psi$, which is true since the possible values of $t$ relies on the size of the binomial.

2) The true value of the parameter does not fall on the boundaries. This is also true since the range of $\psi = (-\infty, 1)$, meaning it does not have set boundaries. 

3) The sample size is significantly large. Our sample size is 170.  


**Bootstrap Confidence Interval**  
The MLE of $\pi$ based on the dataset is $\frac{8}{170}$, we then use this $\pi$ to generate 1000 samples and find $\psi$ for each sample. The value of first and last 2.5% of $\psi$ is Bootstrap percentile 95% confidence interval.

**P-Value Based on the Likelihood Ratio Test Statistic**  
The p-value calculation is based off our null and alternative hypotheses where our null hypothesis assumes that the efficacy is the FDA's required 30% while the alternative suggests that the true rate is not equal to 30%, but rather higher. Thus higher values of $\psi$ support the alternative hypothesis.
$$H_0: \psi_0 = 0.3, \: \: H_1: \psi_0 \neq 0.3$$
Our test statistic is the Likelihood Ratio Test Statistic, denoted:
$$W = 2ln\left(\frac{L(\hat{\psi}_0)}{L({\psi}_0^{null} = 0.3)} \right) \sim \chi^2_1$$

**Empirical P-Value**  
The empirical p-value was calculated based on a large number of randomly generated samples generated with the parameters outlined by the null hypothesis ($\psi_0 = 0.3$). The empirical p-value is equal to the percentage of the generated samples that have a Likelihood Ratio p-value as significant or more significant than the p-value from our original data. 

## 4.2 Bayesian Inference

# 5 Results 
## 5.1 Results from Likelihood Methods  
The maximum-likelihood estimate of the parameter $\psi$ given the data is approximately 0.9506, based on the equations in section 4.1.  
$$\hat{\psi}^{mle}_0= \frac{154}{162} \approx 0.9506$$
And the 95% large sample confidence interval (the Wald Confidence Interval) is:
$$0.9506 \pm z_{0.975} \sqrt{\frac{1}{\frac{8}{0.9506 - 1} - \frac{170}{2-0.9506}}} \approx \big[0.9156, 0.9857\big]$$
Second-order approximation plot\footnote{See appendix 8.x for R code.}:

```{r, warning = F, echo = F}
loglik.binom <- function(psi, x, n){ 
  if (psi >= 1) {
    return(NA)
  }
  pi <- (psi - 1) / (psi - 2)
  return(log(choose(n, x)) + x*log(pi) + (n-x)*log(1-pi))
}

ml.binom <- maxLik2(loglik=loglik.binom,
                    start = 0.3, x = 8, n = 170)

plot(ml.binom) + 
  labs(x = expression(psi),
       title = "Second Order Taylor approximation to Binomial t = 8, n = 170") 
```